Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-18 main.tex
still in IX

\newcommand{\sysS}{\mathsf{S}}
\newcommand{\Deri}{\Phi}
\newcommand{\Proof}{\Pi}

@@

  Now we construct for
  $\Deri$ its propositional encoding $\PE\Deri$ by
  translating every line into its proposional encoding. The instances
  of the rules $\rectif\mfaD$ and $\rectif\mexD$ are replaced two
  steps by:
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlor(\PE{(Ax)}\vlor\PE{(Bx))}}}{
        \vlin{\fequ}{}{\Scons{(y\vlor z)\vlor(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlor\PE{(Ay)})\vlor (z\vlor\PE{(Bz)})}}}}}
  \end{equation*}
  and
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlan\PE{(Ax)}\vlor\PE{(Bx)}}}{
        \vlin{\me}{}{\Scons{(y\vlor z)\vlan(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlan\PE{(Ay)})\vlor (z\vlan\PE{(Bz)})}}}}}
  \end{equation*}
  respectively, where $\rectif\acD$ is a $\acD$ that renames the
  variabes---the atom variable, as well as the first-order variable of
  the same name---as everything is recified, there is no ambiguity
  here.
  This gives us a derivation $\vlderivation{
    \vlde{\set{\wrD,\acD,\rectif\acD,\me,\PE\fequ}}{\PE\Deri}{\PE
      B}{ \vlhy{\PE A}}}$ such that  $\mapof{\PE\Deri}=\mapof\Deri$. It has been
  shown in~\cite{str:07:RTA} that $\mapof{\PE\Deri}$ is a skew
  fibration (see
  also~\cite{hughes:invar,str:ral:tableaux19,str:RR-9048}). Hence,
  $\mapof\Deri$ is a skew fibration.
\end{proof}
\subsection{From Skew Bifibrations to Contraction and Weakening}

@@

\begin{proof}
  By Lemma~\ref{lem:PE} we have $\gA=\graphof{\PE A}$ and
  $\gB=\graphof{\PE B}$.  Let $\vB'\subseteq\vB$ be the image of
  $\phi$, and let $\gB_1$ be the subgraph of $\gB$ induced by
  $\vB'$. Hence, we have two maps
  $\phi''\colon\gA\to\gB_1$ being a surjection and $\phi'\colon
  \gB_1\to\gB$ being an injection that reflects edges. Both, $\phi'$
  and $\phi''$ remain skew bifibrations. Let us first look at $\phi'$.
  Let $\PE B_1$ be the propositional formula obtained from
  $\PE B$ by removing all atoms that are not represented by vertices
  in $\vB'$. Then $\graphof{\PE B_1}=\gB_1$. 
  By \cite[Proposition~7.6.1]{str:07:RTA}, we have a derivation
  $\vlder{\set{\wrD,\fequ}}{\PE\Deri_1}{\PE B}{\PE B_1}$.  A
  subformula of $\PE B$ is called \emph{weak} if it has been
  introduced by a weakening. All subformulas of a weak formula are
  also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  \emph{weak pair} if $B\fequ\Scons{B'\vlor B''}$ for some context
  $\Sconhole$. We now can assume without loss of generality that
  whenever weak subformulas $B'$ and $B''$ for a weak pair, they have
  been introduced by the same instance of $\wlD$ or $\wrD$ in
  $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  brought into this form by simple rule permutations, and then
  collapsing several weakenings into a single one.}  Now observe that
  whenever a weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it
  must also delete all atoms in the scope of the corresponding
  quantifier, because $\phi$ is a fibration on the binding
  graph. Hence, each line in $\PE\Deri_1$ is the propositional
  encoding $\PE P$ of a first-order formula $P$. Thus we have a first
  order derivation $\vlder{\set{\wrD,\fequ}}{\Deri_1}{B}{B_1}$. we now
  have to show that each instance of $\wlD$ or $\wrD$ is indeed a
  correct application in first-order logic. For this we have to
  inspect the cases a weakening happens inside a subformula $x\vlor C$
  or $x\vlan C$ in $\PE\Deri_1$. There are the following cases:
  \begin{equation*}%\small
    \vlinf{}{}{S\cons{x\vlor D\vlor C}}{S\cons{x\vlor C}}
    \qquad
    \vlinf{}{}{S\cons{x\vlan (D\vlor C)}}{S\cons{x\vlan C}}
    \qquad
    \vlinf{}{}{S\cons{(x\vlor D)\vlan C}}{S\cons{x\vlan C}}
  \end{equation*}
  In the first case the weakening happens inside the scope of a
  $\forall$-quantifier, and in the second case inside the scope of a
  $\exists$-quantifier. Both cases are unproblematic in first-order
  logic. However, in the third case an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur, and therefore
  $\Deri_1$ is a valid first-order derivation in $\set{\wD,\fequ}$.

  Let us now look at $\phi''$. Let $\gA_1=\gA\substof\phi$ be the
  graph obtained from $\gA$ by applying $\substof\phi$ to all the
  labels. Note that $\gA_1$ is not neccesarily a fograph, as binders
  might be illegal. But it still is a cograph, and we have a
  surjective skew fibration $\phi''\colon\gA_1\to\gB_1$ that preserves
  the labels. Therefore, by
  \cite[Proposition~7.5]{str:ral:tableaux19}, there is a derivation
  $\vlder{\set{\acD,\me,\PE\fequ}}{\PE\Deri_2}{\PE B_1}{\PE A_1}$,
  where $\PE A_1 =\PE A\substof\phi$ is the result of applying $\substof\phi$
  to~$\PE A$. Note that $\PE A_1 = \PE{(A\substof\phi)}$. To $\Deri_2$ we can apply a permutation argument,
  permuting all instances of $\acD$ up until they either reach the top
  of the derivation or an instance of $\me$ which separates the two
  atoms in the premise. More precisely, we consider the following
  inferene rule
  \begin{equation}
    \vlinf{\acDeq}{}{\Scons{a}}{S_0\cons{S_1\cons{a}\vlor S_2\cons{a}}}
  \end{equation}
  where $S_1\conhole\fequ\conhole\vlor E$ and
  $S_2\conhole\fequ\conhole\vlor F$ and $S\conhole\fequ
  S_0\cons{\conhole\vlor E\vlor F}$ for some formulas $E$ and $F$. The
  rule $\acDeq$ permutes over $\fequ$ and other instances of $\acDeq$,
  and over instances of $\me$ if they occur inside $S_0$ or $S_1$ or
  $S_2$. The only situation in which $\acDeq$ cannot be permuted up is
  the following:
  \begin{equation}
    \label{eq:mac}
    \vlderivation{
      \vlin{\acDeq}{}{\Scons{R\cons{a}\vlan(C\vlor D)}}{
        \vlin{\me}{}{\Scons{(R_1\cons{a}\vlor R_2\cons{a})\vlan(C\vlor D)}}{
          \vlhy{\Scons{(R_1\cons{a}\vlan C)\vlor (R_2\cons{a}\vlor D)}}}}}
  \end{equation}
  Now we translate $\PE\Deri_2$ into a first-order derivation
  $\vlder{\set{\acD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{B_1}{A\substof\phi}$
  by ``decoding'' every line in $\PE\Deri_2$. I.e., for each line $Q$
  in $\PE\Deri_2$ we construct a formula $P$ with $\PE P\fequ Q$.  We
  proceed by induction on the height of $\PE\Deri_2$, beginning at the
  bottom. We already constructed the formula~$B_1$. Now consider the bottommost rule in $\PE\Deri_2$. If it is a $\fequ$, then we trivially can go to the next step. If it is a $\acD$ duplicating an atom that is not a $x\in\VAR$, then this $\acD$ is also a valid inference step in the first-order system. Similarly, any instance of $\me$ that is not in a situation as in~\eqref{eq:mac} can be i
\end{proof}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\newcommand{\sysS}{\mathsf{S}}
\newcommand{\Deri}{\Phi}
\newcommand{\DDeri}{\Psi}
\newcommand{\Proof}{\Pi}

@@

  Now we construct for
  $\Deri$ its propositional encoding $\PE\Deri$ by
  translating every line into its proposional encoding. The instances
  of the rules $\rectif\mfaD$ and $\rectif\mexD$ are replaced two
  steps by:
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlor(\PE{(Ax)}\vlor\PE{(Bx))}}}{
        \vlin{\fequ}{}{\Scons{(y\vlor z)\vlor(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlor\PE{(Ay)})\vlor (z\vlor\PE{(Bz)})}}}}}
  \end{equation*}
  and
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlan\PE{(Ax)}\vlor\PE{(Bx)}}}{
        \vlin{\me}{}{\Scons{(y\vlor z)\vlan(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlan\PE{(Ay)})\vlor (z\vlan\PE{(Bz)})}}}}}
  \end{equation*}
  respectively, where $\rectif\acD$ is a $\acD$ that renames the
  variabes---the atom variable, as well as the first-order variable of
  the same name---as everything is recified, there is no ambiguity
  here.
  This gives us a derivation $\vlderivation{
    \vlde{\set{\wrD,\acD,\rectif\acD,\me,\PE\fequ}}{\PE\Deri}{\PE
      B}{ \vlhy{\PE A}}}$ such that  $\mapof{\PE\Deri}=\mapof\Deri$. It has been
  shown in~\cite{str:07:RTA} that $\mapof{\PE\Deri}$ is a skew
  fibration (see
  also~\cite{hughes:invar,str:ral:tableaux19,str:RR-9048}). Hence,
  $\mapof\Deri$ is a skew fibration.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{From Skew Bifibrations to Contraction and Weakening}

@@

In the proof of this lemma, we make use of the following concept: Let
$\vlder{\sysS}{\DDeri}{Q}{P}$ be a derivation where $Q=\PE D$ is the
propostional encoding of some formula~$D$. We say that $\DDeri$ can be
\bfit{lifted} to $\sysS'$ if there is a formula $C$ such that
$P\fequ\PE C$ and there is a derivation
$\vlder{\sysS'}{\DDeri'}{D}{C}$.

\begin{proof}[Proof of Lemma~\ref{skew->cw}]
  By Lemma~\ref{lem:PE} we have $\gA=\graphof{\PE A}$ and
  $\gB=\graphof{\PE B}$.  Let $\vB'\subseteq\vB$ be the image of
  $\phi$, and let $\gB_1$ be the subgraph of $\gB$ induced by
  $\vB'$. Hence, we have two maps $\phi''\colon\gA\to\gB_1$ being a
  surjection and $\phi'\colon \gB_1\to\gB$ being an injection that
  reflects edges. Both, $\phi'$ and $\phi''$ remain skew
  bifibrations. Let us first look at $\phi'$.  Let $\PE B_1$ be the
  propositional formula obtained from $\PE B$ by removing all atoms
  that are not represented by vertices in $\vB'$. Then $\graphof{\PE
    B_1}=\gB_1$.  By \cite[Proposition~7.6.1]{str:07:RTA}, we have a
  derivation $\vlder{\set{\wrD,\fequ}}{\PE\Deri_1}{\PE B}{\PE B_1}$.
  A subformula of $\PE B$ is called \emph{weak} if it has been
  introduced by a weakening. All subformulas of a weak formula are
  also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  \emph{weak pair} if $B\fequ\Scons{B'\vlor B''}$ for some context
  $\Sconhole$. We now can assume without loss of generality that
  whenever weak subformulas $B'$ and $B''$ for a weak pair, they have
  been introduced by the same instance of $\wlD$ or $\wrD$ in
  $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  brought into this form by simple rule permutations, and then
  collapsing several weakenings into a single one.}  Now we show that
  $\PE\Deri_1$ can be lifted. For this, observe that whenever a
  weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it must also
  delete all atoms in the scope of the corresponding quantifier,
  because $\phi$ is a fibration on the binding graph. Hence, each line
  in $\PE\Deri_1$ is the propositional encoding $\PE P$ of a
  first-order formula $P$.  We now have to show
  that each instance of $\wrD$ is indeed a correct
  application in first-order logic. For this we have to inspect the
  cases a weakening happens inside a subformula $x\vlor C$ or $x\vlan
  C$ in $\PE\Deri_1$. There are the following cases:
  \begin{equation*}%\small
    \vlinf{}{}{S\cons{x\vlor D\vlor C}}{S\cons{x\vlor C}}
    \qquad
    \vlinf{}{}{S\cons{x\vlan (D\vlor C)}}{S\cons{x\vlan C}}
    \qquad
    \vlinf{}{}{S\cons{(x\vlor D)\vlan C}}{S\cons{x\vlan C}}
  \end{equation*}
  In the first case the weakening happens inside the scope of a
  $\forall$-quantifier, and in the second case inside the scope of a
  $\exists$-quantifier. Both cases are unproblematic in first-order
  logic. However, in the third case an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur.
  Thus we have a first order derivation
  $\vlder{\set{\wrD,\fequ}}{\Deri_1}{B}{B_1}$.

  Let us now look at $\phi''$. Let $\gA_1=\gA\substof\phi$ be the
  graph obtained from $\gA$ by applying $\substof\phi$ to all the
  labels. Note that $\gA_1$ is not neccesarily a fograph, as binders
  might be illegal. But it still is a cograph, and we have a
  surjective skew fibration $\phi''\colon\gA_1\to\gB_1$ that preserves
  the labels. Therefore, by
  \cite[Proposition~7.5]{str:ral:tableaux19}, there is a derivation
  $\vlder{\set{\acD,\me,\PE\fequ}}{\PE\Deri_2}{\PE B_1}{\PE A_1}$,
  where $\PE A_1 =\PE A\substof\phi$ is the result of applying $\substof\phi$
  to~$\PE A$. Note that $\PE A_1 = \PE{(A\substof\phi)}$. To $\Deri_2$ we can apply a permutation argument,
  permuting all instances of $\acD$, that duplicate an atom $x\in\VAR$, up until they either reach the top
  of the derivation or an instance of $\me$ which separates the two
  atoms in the premise. More precisely, we consider the following
  inferene rule
  \begin{equation}
    \vlinf{\acDeq}{}{\Scons{x}}{S_0\cons{S_1\cons{x}\vlor S_2\cons{x}}}
  \end{equation}
  where $S_1\conhole\fequ\conhole\vlor E$ and
  $S_2\conhole\fequ\conhole\vlor F$ and $S\conhole\fequ
  S_0\cons{\conhole\vlor E\vlor F}$ for some formulas $E$ and $F$, where $E$ or $F$ or both might be empty. The
  rule $\acDeq$ permutes over $\fequ$, $\acD$, and other instances of $\acDeq$,
  and over instances of $\me$ if they occur inside $S_0$ or $S_1$ or
  $S_2$. The only situation in which $\acDeq$ cannot be permuted up is
  the following:
  \begin{equation}
    \label{eq:mac}
    \vlderivation{
      \vlin{\acDeq}{}{\Scons{R\cons{x}\vlan(C\vlor D)}}{
        \vlin{\me}{}{\Scons{(R_1\cons{x}\vlor R_2\cons{x})\vlan(C\vlor D)}}{
          \vlhy{\Scons{(R_1\cons{x}\vlan C)\vlor (R_2\cons{x}\vlor D)}}}}}
  \end{equation}
  Now we show that $\PE\Deri_2$ can be lifted to
  $\set{\acD,\me,\mfaD,\mexD,\fequ}$.  We proceed by induction on the
  height of $\PE\Deri_2$, beginning at the bottom, as we already
  constructed the formula~$B_1$.  Every instance of $\fequ$ or $\acD$
  (duplicating an atom that is not a $x\in\VAR$), can trivially be
  lifted. Similarly, an instance of $\me$, that is above an instance
  of $\acDeq$ as indicated in~\eqref{eq:mac}, can be lifted as the
  premise must also be a proposional encoding and the $\me$-inference
  remains correct in the first-order case.  It remains to consider a
  situation~\eqref{eq:mac}. Observe that we must have
  $R_1\cons{x}\fequ x\vlor E$ for some $E$ and $R_2\cons{x}\fequ
  x\vlor F$ for some $F$ with $R\cons{x}\fequ x\vlor E\vlor
  F$. Otherwise, the application of $\acDeq$ would not be correct.
  We have the following four cases: 
  \begin{itemize}
  \item   $E$ and $F$ are both empty: Then~\eqref{eq:mac} is
    \begin{equation*}
    \vlderivation{
      \vlin{\acDeq}{}{\Scons{x\vlan(C\vlor D)}}{
        \vlin{\me}{}{\Scons{(x\vlor x)\vlan(C\vlor D)}}{
          \vlhy{\Scons{(x\vlan C)\vlor (x\vlor D)}}}}}
    \end{equation*}
    which can be lifted immediately to
    \begin{equation*}
      \vlinf{\mexD}{}{\Scons{\exists x.(C\vlor D)}}{\Scons{(\exists x.C)\vlor(\exists x.D)}}
    \end{equation*}
  \item $E$ and $F$ are both non-empty: Then~\eqref{eq:mac} is (modulo
    omitted applications of $\fequ$):
    \begin{equation*}
    \vlderivation{
      \vlin{\acDeq}{}{\Scons{(x\vlor E\vlor F)\vlan(C\vlor D)}}{
        \vlin{\me}{}{\Scons{((x\vlor E)\vlor (x\vlor F))\vlan(C\vlor D)}}{
          \vlhy{\Scons{((x\vlor E)\vlan C)\vlor ((x\vlor F)\vlor D)}}}}}
    \end{equation*}
    which can be lifted to
    \begin{equation*}
    \vlderivation{
      \vlin{\mfaD}{}{\Scons{(\forall x.(E\vlor F))\vlan(C\vlor D)}}{
        \vlin{\me}{}{\Scons{((\forall x. E)\vlor (\forall x. F))\vlan(C\vlor D)}}{
          \vlhy{\Scons{((\forall x. E)\vlan C)\vlor ((\forall x. F)\vlor D)}}}}}
    \end{equation*}
  \item $E$ is empty and $F$ is not: Then~\eqref{eq:mac} becomes
    \begin{equation*}
    \vlderivation{
      \vlin{\acDeq}{}{\Scons{(x\vlor F)\vlan(C\vlor D)}}{
        \vlin{\me}{}{\Scons{(x\vlor (x\vlor F))\vlan(C\vlor D)}}{
          \vlhy{\Scons{(x\vlan C)\vlor ((x\vlor F)\vlor D)}}}}}
    \end{equation*}
    The conclusion is the propositional encoding of $\Scons{(\forall
      x. F)\vlan(C\vlor D)}$ and the premise is the propositional
    encoding of $\Scons{(\exists x.C)\vlor ((\forall x. F)\vlor
      D)}$. Also note that no $\me$-instance can break up the
    conjunction in $x\vlan C$ in the premise. Hence, $\phi$ maps an
    existential to a universal, which is ruled out by the
    definition. Hence, this case cannot occur.
  \item $E$ is non-empty and $F$ is empty: This case is similar to the
    previous one.
  Finally, we need to consider the instances of $\acDeq$ that are at the top of the derivation.
